---
title: 'Coursera Reproducible Research Course Project 2: Severe Weather Events'
author: "J Cecire"
date: "9/11/2021"
output:
  pdf_document: default
  html_document: default
---
## Synopsis
This project investigates the severe weather events in the US based on the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. (For a detailled description of the NOAA storm database see: https://www.ncdc.noaa.gov/stormevents/)
Using the NOAA storm database this analysis clearly shows that tornados have the most harmful impact on people’s health as they resulted in the highest number of fatalities and injures. This analysis also revealed that floods caused the most property damage in terms of price and droughts caused the crop damage in terms of price.

## Questions
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

## Loading Packages 
```{r setup, echo=TRUE}
library(dplyr)
library(knitr)
```

## Loading Data
```{r, echo=TRUE}
stormData <- read.csv("./repdata-data-StormData.csv.bz2")
summary(stormData)
names(stormData)
str(stormData)
```

## Wranggling Data

### Extracting needed Variables
Since for the rest of this analysis we will only be focusing on events that either harm the populations health or impact the economy we can wrangle and consolodate the data getting rid of unnecessary columns. The columns we will focus on are as follows:

EVTYPE: Event Type (Tornados, Flood, ….)

FATALITIES: Number of Fatalities

INJURIES: Number of Injuries

PROPDMG: Property Damage

PROPDMGEXP: Units for Property Damage (magnitudes - K,M,B)

CROPDMG: Crop Damage

CROPDMGEXP: Units for Crop Damage (magnitudes - K,M,B)
```{r, echo=TRUE}
keyVars <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
storm <- stormData[keyVars]
dim(storm)
names(storm)
summary(storm)
str(storm)
```

### Making the PROPDMGEXP and CROPDMGEXP columns cleaner so they can be used to calculate property and crop cost.
- Using this document (https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) we can tell what each unique value stands for in the PROPDMGEXP and CROPDMGEXP columns and how they relate to the PROPDMG and CROPDMG variables. So, we must first recode the information in columns PROPDMGEXP and CROPDMGEXP and then refactor them. 
```{r, echo=TRUE}
unique(storm$PROPDMGEXP)
```
- Recoding and refactoring of variable PROPDMGEXP
```{r, echo=TRUE}
storm$PROPDMGEXP <- recode(storm$PROPDMGEXP, 
                              "K" = 10^3, "M" = 10^6, " " = 1, "B" = 10^9, "m" =10^6, "+" = 0, '0' = 1, '5' = 10^5, '6' =10^6, "?" = 0, '4' = 10^4, '2' = 10^2, '3' = 10^3, "h" = 10^2, '7' = 10^7, "H" = 10^2, "-" = 0, '1' = 10, '8' = 10^8)

storm$PROPDMGEXP <- as.numeric(as.character(storm$PROPDMGEXP))
storm$PROPDMGTOTAL <- (storm$PROPDMG * storm$PROPDMGEXP)/1000000000
```

```{r, echo=TRUE}
unique(storm$CROPDMGEXP)
```
- Recoding and refactoring of variable CROPDMGEXP
```{r, echo=TRUE}
storm$CROPDMGEXP <- recode(storm$CROPDMGEXP, 
                              " " = 1,"M" = 10^6, "K" = 10^3, "m" = 10^6, "B" = 10^9, "?" = 0, "0" = 1, "k" = 10^3,"2" = 10^2)

storm$CROPDMGEXP <- as.numeric(as.character(storm$CROPDMGEXP))
storm$CROPDMGTOTAL <- (storm$CROPDMG * storm$CROPDMGEXP)/1000000000
```

## Which type of events are most harmful to human health?
- Since there are 985 different event types we are only going to look at the top 10 most fatal events and top 10 injury events.

### Fatalities 
```{r, echo=TRUE}
stormFatalities <- aggregate(FATALITIES ~ EVTYPE, data = storm,  FUN="sum")
dim(stormFatalities)
names(stormFatalities)
summary(stormFatalities)
str(stormFatalities)
```

- Ordering the top 10 Weather events by number of fatalities
```{r, echo=TRUE}
top10FatalEvents <- stormFatalities[order(-stormFatalities$FATALITIES), ][1:10, ]
top10FatalEvents

```
### Plot Fatalities
```{r, echo=TRUE}
par(mfrow = c(1,1), mar = c(10, 4, 4, 2), mgp = c(3, 1, 0), cex = 0.7)
barplot(top10FatalEvents$FATALITIES, names.arg = top10FatalEvents$EVTYPE, las = 3, 
        main = "Top 10 Fatal Events", 
        ylab = "Number of Fatalities",
        ylim = range(0,6000),
        col = c("red"))

```


### Injuries
```{r, echo=TRUE}
stormInjuries <- aggregate(INJURIES ~ EVTYPE, data = storm,  FUN="sum")
dim(stormInjuries)
names(stormInjuries)
summary(stormInjuries)
str(stormInjuries)
```

- Ordering the top 10 Weather events by number of injuries
```{r, echo=TRUE}
top10InjuryEvents <- stormInjuries[order(-stormInjuries$INJURIES), ][1:10, ]
top10InjuryEvents
```

### Plot Injuries
```{r, echo=TRUE}
options(scipen = 100)
par(mfrow = c(1,1), mar = c(11, 6, 4, 2), mgp = c(3, 1, 0), cex = 0.7)
barplot(top10InjuryEvents$INJURIES, names.arg = top10InjuryEvents$EVTYPE, las = 3, 
        main = "Top 10 Injury Events", 
        ylab = "Number of Injuries",
        ylim = range(0,100000),
        col = c("Orange"))

```

## Which type of events have the greatest economic consequences? 
- Similar to the human health questions, since there are 985 different event types we are only going to look at the top 10 most damaging property events and top 10 most damaging crop events.

### Property Damage
```{r, echo=TRUE}
stormPropDmg <- aggregate(PROPDMGTOTAL ~ EVTYPE, data = storm,  FUN="sum")
dim(stormPropDmg)
names(stormPropDmg)
summary(stormPropDmg)
str(stormPropDmg)
```

- Ordering the top 10 Weather events by property damage amounts
```{r, echo=TRUE}
top10PropDmgEvents <- stormPropDmg[order(-stormPropDmg$PROPDMGTOTAL), ][1:10, ]
top10PropDmgEvents

```

### Plot Property Damage
```{r, echo=TRUE}
par(mfrow = c(1,1), mar = c(12, 6, 4, 2), mgp = c(3, 1, 0), cex = 0.7)
barplot(top10PropDmgEvents$PROPDMGTOTAL, names.arg = top10PropDmgEvents$EVTYPE, las = 3, 
        main = "Top 10 Property Damage Events", 
        ylab = "Property Damage Values (in Billions)",
        ylim = range(0,200),
        col = c("Black"))

```

### Crop Damage
```{r, echo=TRUE}
stormCropDmg <- aggregate(CROPDMGTOTAL ~ EVTYPE, data = storm,  FUN="sum")
dim(stormCropDmg)
names(stormCropDmg)
summary(stormCropDmg)
str(stormCropDmg)
```

- Ordering the top 10 Weather events by crop damage amounts
```{r, echo=TRUE}
top10CropDmgEvents <- stormCropDmg[order(-stormCropDmg$CROPDMGTOTAL), ][1:10, ]
top10CropDmgEvents

```

### Plot Crop Damage
```{r, echo=TRUE}
par(mfrow = c(1,1), mar = c(12, 6, 4, 2), mgp = c(3, 1, 0), cex = 0.7)
barplot(top10CropDmgEvents$CROPDMGTOTAL, names.arg = top10CropDmgEvents$EVTYPE, las = 3, 
        main = "Top 10 Crop Damage Events", 
        ylab = "Crop Damage Values (in Billions)",
        ylim = range(0,20),
        col = c("Green"))

```

